Fractional Types

### 2013-02-07 frac.tex
relational semantics confirmed opeerational semantics deleted

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! langRevT = \Pi^{o}
%! langRevTF = \Pi^{o/}

@@

{{langRev}} models isomorphisms of values rather well.  However, purposefully,
it has no notion of \emph{input} or \emph{output}; more precisely, because
of strict preservation of information, these two concepts coincide.  This
is the root of reversibility.

But if we want to model functions of any flavor, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position in a type to possess negative
information, as it is really a request for information.  Since information
is logarithmic, this means that a request for information should behave
like a \emph{fractional type}.

@@

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in the category of finite sets and relations. This category is
an example of a \emph{biproduct dagger compact closed category} which are
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} The category intuitively supports
higher-order relations as any relation between $A$ and $B$ can be represented
as a subset of {{A * B}}, i.e., as a value. More generally, the two main
features of these biproduct dagger compact closed categories are that they
support an adjoint for each morphism (to model reversibility) and a notion of
higher-order relation (built from the dagger which is modeled by our
fractional types).

To interpret {{langRevTF}} in this category, we associate a finite set
$\llbracket b \rrbracket$ with each type $b$ and a relation $\llbracket c
\rrbracket$ with each combinator $c$. The denotation of types is
straighforward: the denotation of {{1/b}} is an isomorphic copy of the
denotation of {{b}}.

@@

@@

Perhaps get rid of the semantics below and replace by a semantics which maps
each combinator to a relation. This is probably clearer than the operational
style.

@@

present formal semantics

model using vector spaces; category of finite sets and relations; homotopy
equivalence???

The operational semantics of {{langRevTF}} is given using two mutually
recursive interpreters: one going forward and one going backwards. Unlike the
case for {{langRev}}, each combinator denotes a \emph{relation} and hence the
result of applying a combinator to a value generally returns a \emph{set} of
values.

\begin{definition}[Operational semantics for {{langRevTF}}]
\label{def:operational-langRevTF}
Given a program {{c : b1 <-> b2}} in {{langRev}}, we can run it in the
forward direction by supplying it with a value {{ v1 : b1 }} and in
the backwards direction by supplying it with a value {{v2 : b2}}.

\paragraph*{Forward Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalF swap+ (left v) &=& {[ right v ]}
% evalF swap+ (right v) &=& {[ left v ]}
% evalF assocl+ (left v) &=& {[ left (left v) ]}
% evalF assocl+ (right (left v)) &=& {[ left (right v) ]}
% evalF assocl+ (right (right v)) &=& {[ right v ]}
% evalF assocr+ (left (left v)) &=& {[ left v ]}
% evalF assocr+ (left (right v)) &=& {[ right (left v) ]}
% evalF assocr+ (right v) &=& {[ right (right v) ]}
% evalF unite ((), v) &=& {[ v ]}
% evalF uniti v &=& {[ ((), v) ]}
% evalF swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalF assocl* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalF assocr* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalF dist (left v1, v3) &=& {[ left (v1, v3) ]}
% evalF dist (right v2, v3) &=& {[ right (v2, v3) ]}
% evalF factor (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalF factor (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalF id v &=& {[ v ]}
% evalF (sym c) v &=& evalB c v
% evalF (c1(;)c2) v &=& Union ~{[ evalF c2 v' ~|~ v' in (evalF c1 v) ]}
% evalF (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalF c1 v1) ]}
% evalF (c1 (+) c2) (right v2) &=& {[ left v' ~|~ v' in (evalF c2 v2) ]}
% evalF (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalF c1 v1), v4 in (evalF c2 v2) ]}
% evalF ~eta*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
% evalF ~eps*_b (1/v,v) &=& {[ () ]} 
% evalF ~eps*_b (1/v1,v2) &=& emptyset, where v1 /= v2
\paragraph*{Backwards Evaluator:}

%subcode{opsem} include main
%! columnStyle = rcl
% evalB swap+ (left v) &=& {[ right v ]}
% evalB swap+ (right v) &=& {[ left v ]}
% evalB assocl+ (left (left v)) &=& {[ left v ]}
% evalB assocl+ (left (right v)) &=& {[ right (left v) ]}
% evalB assocl+ (right v) &=& {[ right (right v) ]}
% evalB assocr+ (left v) &=& {[ left (left v) ]}
% evalB assocr+ (right (left v)) &=& {[ left (right v) ]}
% evalB assocr+ (right (right v)) &=& {[ right v ]}
% evalB unite v &=& {[ ((), v) ]}
% evalB uniti ((), v) &=& {[ v ]}
% evalB swap* (v1, v2) &=& {[ (v2, v1) ]}
% evalB assocl* ((v1, v2), v3) &=& {[ (v1, (v2, v3)) ]}
% evalB assocr* (v1, (v2, v3)) &=& {[ ((v1, v2), v3) ]}
% evalB dist (left (v1, v3)) &=& {[ (left v1, v3) ]}
% evalB dist (right (v2, v3)) &=& {[ (right v2, v3) ]}
% evalB factor (left v1, v3) &=& {[ left (v1, v3) ]}
% evalB factor (right v2, v3) &=& {[ right (v2, v3) ]}
% evalB id v &=& {[ v ]}
% evalB (sym c) v &=& evalF c v
% evalB (c1(;)c2) v &=& Union ~{[ evalB c1 v' ~|~ v' in (evalB c2 v) ]}
% evalB (c1 (+) c2) (left v1) &=& {[ left v' ~|~ v' in (evalB c1 v1) ]}
% evalB (c1 (+) c2) (right v2) &=& {[ right v' ~|~ v' in (evalB c2 v2) ]}
% evalB (c1 (*) c2) (v1,v2) &=& {[ (v3,v4) ~|~ v3 in (evalB c1 v1), v4 in (evalB c2 v2) ]}
% evalB ~eta*_b (1/v,v) &=& {[ () ]} 
% evalB ~eta*_b (1/v1,v2) &=& emptyset, where v1 /= v2
% evalB ~eps*_b () &=& Union_{v in [[b]]} ~{[ (1/v,v) ]}
\end{definition}

@@

What is interesting about this isomorphism is that it holds for all
types {{b}}, and in particular we have {{0 <-> 1/(1/0)}}.

@@

\begin{example}
A more fundamental problem occurs if we attempt to apply the trace
operator to cancel the empty type. Consider a combinator:

{{c : 0 * 1 <-> 0 * bool}}

{{c = distrib0 (;) factor0}}

\noindent Using the multiplicative trace, we can construct the combinator
{{trace* c : 1 <-> bool}} which relates two types of different sizes! Fortunately,
because {{0}} is the empty type, 
when attempting to evaluate such a combinator, it is impossible to find a value
{{v : 0}} to satisfy the constraint imposed by {{trace*}} and hence 
we would also expect the evaluation of such a combinator to produce no value.
\end{example}

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! [^ = \ulcorner
%! ^] = \urcorner
%! [v = \llcorner
%! v] = \lrcorner
%| [[ = \llbracket
%| ]] = \rrbracket
%! eta* = \eta^{\times}
%! eps* = \epsilon^{\times}
%! Union = \bigcup
%! in = \in
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |-->let = \mapsto_{let}
%! |--> = \mapsto
%! <--| = \mapsfrom
%! |- = \vdash
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! ::= = ::=
%! /= = \neq
%! vi = v_i
%! di = d_i
%! si = s_i
%! sj = s_j
%! F = \texttt{F}
%! T = \texttt{T}
%! forall = \forall
%! exists = \exists
%! empty = \emptyset
%! eta = \eta
%! where = \textbf{where}
%! epsilon = \varepsilon
%! least = \phi
%! loop+ = loop_{+}
%! loop* = loop_{\times}
%! CatC = {\mathcal C}
%! CatA = {\mathcal A}
%! gamma = \gamma
%! {[ = \{
%! ]} = \}
%! elem = \in
%! dagger = ^\dagger
%! alpha = \alpha
%! beta = \beta
%! rho = \rho
%! @@ = \mu
%! @ = \,@\,
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}_{+}
%! bullet = \bullet
%! * = \times
%! langRevT = \Pi^{o}
%! langRevTF = \Pi^{/}

@@

The language {{langRev}} models isomorphisms of values rather well.  However,
purposefully, it has no notion of \emph{input} or \emph{output}; more
precisely, because of strict preservation of information, these two concepts
coincide.  This is the root of reversibility.

But if we want to model functions of any flavor, we need to differentiate
between these notions.  Our idea, inspired by computational dualities and
polarity, is to introduce a \emph{formal dual} for our types.  We thus
consider a type in a negative position to possess negative information, as it
is really a request for information.  Since information is logarithmic, this
means that a request for information should behave like a \emph{fractional
  type}.

@@

Even though the categorical semantics of {{langRevTF}} is not the main focus
of this paper, we find it helpful to briefly outline an interpretation of
{{langRevTF}} in the category of finite sets and relations. This category is
an example of a \emph{biproduct dagger compact closed category} which are
categories introduced by Abramsky and
Coecke~\cite{Abramsky:2004:CSQ:1018438.1021878} and by
Selinger~\cite{Selinger:2007:DCC:1229185.1229207} to model quantum
computation.\footnote{The other common example of these categories is that of
  finite dimensional Hilbert spaces.} The two main features of these
biproduct dagger compact closed categories are that they support an adjoint
for each morphism (to model reversibility) and that they support higher-order
relation (built from the categorical dagger which is modeled by our
fractional types). The latter is intuitively clear as any relation between
$A$ and $B$ can be represented as a subset of {{A * B}}, i.e., as a
value.

To interpret {{langRevTF}} in this category, we associate a finite set
{{ [[b]] }} with each type $b$ and a relation {{ [[c]] }} over
{{ [[b1]] }} and {{ [[b2]] }} for each combinator {{ c : b1 <-> b2 }}.
The denotation of types is straighforward: the denotation of {{1/b}} 
is an isomorphic copy of the denotation of {{b}}.

@@

The denotation of combinators as relations is given below. We specify
relations using a deductive system whose judgments are of the form 
{{ v1 ~~c~~ v2 }} indicating that the pair {{(v1,v2)}} is in the relation 
denoted by {{c}}.

\begin{definition}[Relational semantics]
\label{def:relational-langRevTF}
Each combinator {{c : b1 <-> b2}} in {{langRevTF}} denotes a relation
as specified below.

%subcode{proof} include main
%@ ~
%@@ (left v) ~~swap+~~ (right v)
%
%@ ~
%@@ (right v) ~~swap+~~ (left v)
%----
%@ ~
%@@ (left v) ~~assocl+~~ (left (left v))
%
%@ ~
%@@ (right (left v)) ~~assocl+~~ (left (right v))
%
%@ ~
%@@ (right (right v)) ~~assocl+~~ (right v)
%----
%@ ~
%@@ (left (left v)) ~~assocr+~~ (left v)
%
%@ ~
%@@ (left (right v)) ~~assocr+~~ (right (left v))
%
%@ ~
%@@ (right v) ~~assocr+~~ (right (right v))
%----
%@ ~
%@@ ((), v) ~~unite~~ v
%
%@ ~
%@@ v ~~uniti~~ ((), v)
%----
%@ ~
%@@ (v1, v2) ~~swap*~~ (v2, v1)
%----
%@ ~
%@@ (v1, (v2, v3)) ~~assocl*~~ ((v1, v2), v3)
%
%@ ~
%@@ ((v1, v2), v3) ~~assocr*~~ (v1, (v2, v3))
%----
%@ ~
%@@ (left v1, v3) ~~dist~~ (left (v1, v3))
%
%@ ~
%@@ (right v2, v3) ~~dist~~ (right (v2, v3))
%----
%@ ~
%@@ (left (v1, v3)) ~~factor~~ (left v1, v3)
%
%@ ~
%@@ (right (v2, v3)) ~~factor~~ (right v2, v3)
%----
%@ ~
%@@ v ~~id~~ v
%
%@ v' ~~c~~ v
%@@ v ~~(sym c)~~ v'
%----
%@ v1 ~~c1~~ v2
%@ v2 ~~c2~~ v3
%@@ v1 ~~(c1(;)c2)~~ v3
%
%@ v ~~c1~~ v'
%@@ (left v) ~~(c1 (+) c2)~~ (left v')
%
%@ v ~~c2~~ v'
%@@ (right v) ~~(c1 (+) c2)~~ (right v')
%----
%@ v1 ~~c1~~ v1'
%@ v2 ~~c2~~ v2'
%@@ (v1, v2) ~~(c1 (*) c2)~~ (v1', v2')
%----
%@ v in [[b]]
%@@ () ~~eta*_b~~ (1/v,v)
%
%@ ~
%@@ (1/v,v) ~~eps*_b~~ ()

\newpage

@@

@@

homotopy equivalence???

@@

@@
